Explicit Scale Simulation for analysis of RNA-sequencing with ALDEx2

### 2024-08-22 go3.tex
with Justin's suggestions
Send back to Justin for comment

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{setspace}
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Normalizations are not what you think; Explicit Scale Simulation in ALDEx2},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Normalizations are not what you think; Explicit Scale Simulation
in ALDEx2}
\author{true \and true \and true}
\date{}

\begin{document}
\maketitle
\begin{abstract}
In high-throughput sequencing (HTS) studies, sample-to-sample variation
in sequencing depth is driven by technical factors, and not by variation
in the scale (e.g., total size, microbial load, or total gene
expression) of the underlying biological systems. Typically a
statistical normalization is used to remove unwanted technical variation
in the data or the parameters of the model to enable analyses that are
sensitive to scale; e.g., differential abundance and differential
expression analyses. Recently we showed that all normalizations make
implicit assumptions about the unmeasured system scale and that errors
in these assumptions can lead to dramatic increases in false positive
and false negative rates. Here we describe updates to the ALDEx2 R
package that mitigate these problems by directly modeling uncertainty in
the unmeasured system scale through the use of a \textit{scale model}.
Scale models generalize the idea of normalizations and can be thought of
as explicitly modeling the error in normalization by examining a
distribution over all possible normalizations. Beyond enhancing the
robustness of HTS analyses, the use of scale models within ALDEx2
enhances the transparency and reproducibility of analyses by making
implicit normalizing assumptions an explicit part of the model building
process.
\end{abstract}

@@

\doublespacing
\singlespacing

Biological systems are both predictably variable and stochastic (45) and
current measurement methods that rely on high throughput sequencing fail
to capture all of that variation, particularly variation due to scale
(8, 9). In the absence of external information (such as spike-in probes
(14), cell counting (1), FISH (51) etc) sequencing depth normalisation
methods cannot recapture the scale information (14), and can only
normalize for the technical variation due to sequencing depth.

Many groups have conducted benchmarking studies on different tools and
normalizations used for the analysis of datasets such as
transcriptomes(4, 21, 35, 37, 41) and microbiomes (3, 23, 24, 32, 34,
52). Generically, it is observed that the actual agreement between
analysis methods can be modest, and no single method outperforms all
others in every dataset or type of experiment. That different tools
appear to work more reliably in different datasets from different
sources can be explained as different normalizations being a better fit
to the scale of a particular dataset by chance.

In the analysis of HTS data it is often observed that larger datasets
converge on the majority of parts being significantly different (8, 35,
37), and that different analytic approaches result in different parts
being chosen as significantly different (4, 24, 37, 41). Li et al. (35)
conducted a permutation-based benchmarking study and found that popular
tools performed worse than simple Wilcoxon rank-sum tests in controlling
the FDR when sample sizes became large. They suggested that the presence
of outliers were one of the factors driving this observation. Brooks et
al. (53) suggested that inappropriate choice of benchmarking methods are
also a major contributing factor. From our perspective, the disagreement
between tools can be explained by the partially identifiable nature of
the data; that is, the observed data are consistent with multiple ways
of estimating the parameters (8). Partially identifiable data exhibit
multiple pathologies, chief among them being that that different
analytic approaches can produce different parameter estimates and that
more data produces worse estimates because the additional data increases
the precision of a flawed estimate (54).

Scale simulation is now build into ALDEx2 to address the issue of
partially identifiability, and through this lens the two main root
causes to common HTS data pathologies can be proposed. The first
contributing factor is the observed very low dispersion estimate for
many features that is a by-product of experimental design, sequencing
and normalization. In the Schurch et al. (37) dataset, the data were
from single colonies derived from a single culture. Thus, it is more
accurate to describe the 96 samples as wet-lab technical replicates
rather than independent samples. However, this type of replication
approach is standard in the molecular literature, and would be expected
to result in the very low dispersion that is observed.

In the yeast transcriptome dataset, applying the default scale model
with \(\gamma=0.5\) a large number of transcripts with near 0 dispersion
have had their dispersion increased (Figure 1D), and this results in
modest number of transcripts, 205, being called significantly different
as shown in the volcano plot in Figure 1D (red points). In addition,
there was now a strong concordance between the difference and p-values
(Figure 1D), this is not surprising. In hindsight, what is not obvious
is why the unscaled volcano plot shows such poor correspondence. We
suggest that this can be explained by random fluctuations in the many
variance estimates with very low values, and this is supported by the
plot shown in Figure 2B. Furthermore, overplotting the significant
transcripts identified after adding scale uncertainty on the un-scaled
analysis shows that adding scale uncertainty removes the need for the
dual cutoff. Indeed, adding scale uncertainty reduces the significant
transcripts to a subset of those identified with the dual cutoff that
have the largest effect size. Thus, incorporating scale uncertainty
through the default scale model allows us to determine which variables
are likely to be significant due to sequencing and normalization, and
which are significantly different even with scale uncertainty included.

While the actual scale of the underlying environment is inaccessible
post-sequencing, we can study the sensitivity of the results to the
choice of normalization and thus scale. Varying amounts of scale
uncertainty is incorporated using the \texttt{gamma} parameter that
controls the amount uncertainty added to the CLR mean assumption when we
call either \texttt{aldex()}, or \texttt{aldex.clr()}. The ALDEx2
package contains a sensitivity analysis function,
\texttt{aldex.senAnalysis()}, that can be used to explore the effect of
different amounts of scale uncertainty, and an example is shown in
Supplementary Figure 2 for the yeast transcriptome dataset. Here it is
clear that even tiny amounts of scale preclude the majority of
transcripts from being considered statistically significant. In
practice, we suggest that a \texttt{gamma} parameter between 0.5 and 1
is realistic for most experimental designs, but further work is needed.

The second contributing factor is unacknowledged asymmetry in many
datasets. In the case of asymmetry, the use of a user-specified scale
model can be very useful for otherwise difficult-to-analyze datasets
such as meta-transcriptomes and in-vitro selection datasets where the
majority of features can change. We showed one such example for the
metatranscriptome dataset in Figure 3. Here the dataset was highly
asymmetrical, and the TMM and RLE normalizations could not fully move
all the housekeeping genes to the midline of no difference or the tools
exhibited other pathologies (Supplemental Figure 6). Incorporating
differential scale on a per-group basis moves the mass of the
housekeeping functions towards the midline of no difference and so
affects both Type I and Type II error rates. It is also of note that in
the case of true biological replicates (different individuals) that
adding a modest amount of scale \(\gamma=0.5\) had little effect on the
concordance between the difference between groups and the p-values as
shown in Supplementary Figure 7. Thus, in this dataset the scale
mis-specification was affecting mainly the location of the difference
between groups.

In the metatranscriptome analysis, transcripts that were previously not
classed as differentially abundant were now called as significantly
different, and the housekeeping transcripts move from being
significantly different to not being identified as such. Note the
assumption that housekeeping genes should not generally be included in
differential abundance analysis is implicit in the dual
p-value:fold-change cutoff approach in widespread use. While we
acknowledge that some prior information on which housekeeping
transcripts should not be classed as differentially abundant is needed,
we suggest that this information is widely available and is already used
when performing the gold-standard quantitative PCR test of differential
abundance (55, 56). Thus, the use of this prior knowledge is not unique
to our approach.

Beyond concerns of fidelity and rigor, scale models also enhance the
reproducibility and transparency of HTS analyses. The addition of scale
uncertainty essentially tests the model over a range of normalizations
(8) and so can replace the consensus approach that has been proposed by
some groups (24, 57) with no additional computational overhead. Thus, an
advantage of incorporating scale is that analyses can be made much more
robust such that actual or potential differences in scale can be tested
and accounted for explicitly. This occurs because rather than using an
implicit assumption (e.g., \(\log \mathbf{W}^{\perp}=1/G_{n}\) for the
CLR normalization), scale models can make this an explicit part of the
model building process. While it is beyond the scope of the present
article, we note that there are many ways of building scale models that
enhance the interpretability of the parameters and assumptions and a
detailed description of these points is describe elsewhere (8). In this
work, we simply use the parameter \(\gamma\) in the default scale model
as a term that represents uncertainty in the CLR assumption and note
that larger values of \(\gamma\) correspond to more uncertainty in this
assumption. In the metatranscriptome example and elsewhere (9), we also
show that the ratio of scales between conditions can be used to build a
full scale model when the conditions have dramatically different
underlying scales.

The ALDEx2 R package is readily amenable for incorporating scale
uncertainty. Originally, this tool used the only the Dirichlet
distribution to sample compositional uncertainty, but scale uncertainty
can be added through the use of a scale model with no additional
computational complexity. By default, ALDEx2 samples scale from a
logNormal distribution inspired by the CLR normalization. However, scale
uncertainty can be sampled from any distribution depending on prior
knowledge or preference, using the option to introduce a full informed
scale model that encapsulates both uncertainty and asymmetry in
underlying scale.

In summary, we supply a toolkit that makes incorporating scale simple to
incorporate for any type of HTS dataset. While the underlying scale of
the system is generally inaccessible, the effect of scale on the
analysis outcomes can be modelled and can help explain some of the
underlying biology, and known issues with the analysis of HTS data.
Adding scale information to the analysis allows for more robust
inference because the features that are sensitive to scale can be
identified and their impact on conclusions weighted accordingly.
Additionally, the use of user-defined scale models permits difficult to
analyze datasets to be examined in a robust and principled manner even
when the majority of features are asymmetrically distributed or
expressed (or both) in the groups (46). Thus, reporting scale
uncertainty should become a standard practice in the analysis of HTS
datasets as a way to identify which features are most robust to
differences in the underlying system.

@@

\phantomsection\label{refs}
\begin{CSLReferences}{1}{1}
\bibitem[\citeproctext]{ref-Vandeputte:2017aa}
1. Vandeputte,D., Kathagen,G., D'hoe,K., Vieira-Silva,S.,
Valles-Colomer,M., Sabino,J., Wang,J., Tito,R.Y., De Commer,L.,
Darzi,Y., \emph{et al.} (2017)
\href{https://doi.org/10.1038/nature24460}{Quantitative microbiome
profiling links gut community variation to microbial load}.
\emph{Nature}, \textbf{551}, 507--511.

\bibitem[\citeproctext]{ref-Props:2017aa}
2. Props,R., Kerckhof,F.-M., Rubbens,P., De Vrieze,J., Hernandez
Sanabria,E., Waegeman,W., Monsieurs,P., Hammes,F. and Boon,N. (2017)
\href{https://doi.org/10.1038/ismej.2016.117}{Absolute quantification of
microbial taxon abundances}. \emph{ISME J}, \textbf{11}, 584--587.

\bibitem[\citeproctext]{ref-Thorsen:2016aa}
3. Thorsen,J., Brejnrod,A., Mortensen,M., Rasmussen,M.A., Stokholm,J.,
Al-Soud,W.A., Sørensen,S., Bisgaard,H. and Waage,J. (2016)
\href{https://doi.org/10.1186/s40168-016-0208-8}{Large-scale
benchmarking reveals false discoveries and count transformation
sensitivity in 16{S} r{RNA} gene amplicon data analysis methods used in
microbiome studies}. \emph{Microbiome}, \textbf{4}, 62.

\bibitem[\citeproctext]{ref-Quinn:2018aa}
4. Quinn,T.P., Crowley,T.M. and Richardson,M.F. (2018)
\href{https://doi.org/10.1186/s12859-018-2261-8}{Benchmarking
differential expression analysis tools for RNA-seq: Normalization-based
vs. Log-ratio transformation-based methods}. \emph{BMC Bioinformatics},
\textbf{19}, 274.

\bibitem[\citeproctext]{ref-Zhang:2009aa}
5. Zhang,S. and Cao,J. (2009)
\href{https://doi.org/10.1186/1471-2105-10-402}{A close examination of
double filtering with fold change and t test in microarray analysis}.
\emph{BMC Bioinformatics}, \textbf{10}, 402.

\bibitem[\citeproctext]{ref-Ebrahimpoor:2021aa}
6. Ebrahimpoor,M. and Goeman,J.J. (2021)
\href{https://doi.org/10.1093/bib/bbab053}{Inflated false discovery rate
due to volcano plots: Problem and solutions}. \emph{Brief Bioinform},
\textbf{22}.

\bibitem[\citeproctext]{ref-Wu2021}
7. Wu,J.R., Macklaim,J.M., Genge,B.L. and Gloor,G.B. (2021)
\href{https://doi.org/10.1007/978-3-030-71175-7_17}{Finding the centre:
Compositional asymmetry in high-throughput sequencing datasets}. In
Filzmoser,P., Hron,K., Martìn-Fernàndez,J.A., Palarea-Albaladejo,J.
(eds), \emph{Advances in compositional data analysis: Festschrift in
honour of vera pawlowsky-glahn}. Springer International Publishing,
Cham, pp. 329--346.

\bibitem[\citeproctext]{ref-nixon2023scale}
8. Nixon,M.P., Letourneau,J., David,L.A., Lazar,N.A., Mukherjee,S. and
Silverman,J.D. (2023) \href{https://arxiv.org/abs/2201.03616}{Scale
reliant inference}.

\bibitem[\citeproctext]{ref-Nixon2024B}
9. Nixon,M.P., Gloor,G.B. and Silverman,J.D. (2024) Beyond
normalization: Incorporating scale uncertainty in microbiome and gene
expression analysis. \emph{bioRxiv},
\href{https://doi.org/10.1101/2024.04.01.587602}{10.1101/2024.04.01.587602}.

\bibitem[\citeproctext]{ref-lovell:2011}
10. Lovell,D., Müller,W., Taylor,J., Zwart,A. and Helliwell,C. (2011)
Proportions, percentages, ppm: Do the molecular biosciences treat
compositional data right? In Pawlowsky-Glahn,V., Buccianti,A. (eds),
\emph{Compositional Data Analysis: Theory and Applications}. John Wiley;
Sons New York, NY, London, pp. 193--207.

\bibitem[\citeproctext]{ref-Quinn:2019aa}
11. Quinn,T.P., Erb,I., Gloor,G., Notredame,C., Richardson,M.F. and
Crowley,T.M. (2019) \href{https://doi.org/10.1093/gigascience/giz107}{A
field guide for the compositional analysis of any-omics data}.
\emph{Gigascience}, \textbf{8}.

\bibitem[\citeproctext]{ref-Lovell:2015}
12. Lovell,D., Pawlowsky-Glahn,V., Egozcue,J.J., Marguerat,S. and
Bähler,J. (2015)
\href{https://doi.org/10.1371/journal.pcbi.1004075}{Proportionality: A
valid alternative to correlation for relative data}. \emph{PLoS Comput
Biol}, \textbf{11}, e1004075.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{setspace}
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Explicit Scale Simulation for analysis of RNA-sequencing with ALDEx2},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Explicit Scale Simulation for analysis of RNA-sequencing with
ALDEx2}
\author{true \and true \and true}
\date{}

\begin{document}
\maketitle
\begin{abstract}
In high-throughput sequencing (HTS) studies, sample-to-sample variation
in sequencing depth is driven by technical factors, and not by variation
in the scale (e.g., total size, microbial load, or total mRNA
expression) of the underlying biological systems. Typically a
statistical normalization is used to remove unwanted technical variation
in the data or the parameters of the model to enable analyses that are
sensitive to scale; e.g., differential abundance and differential
expression analyses. We recently showed that all normalizations make
implicit assumptions about the unmeasured system scale and that errors
in these assumptions can dramatically increase false positive and false
negative rates. We demonstrated that these errors can be mitigated by
accounting for uncertainty about scale using a \emph{scale model}, which
we integrated into the ALDEx2 R package. This article provides new
insights into those methods, focusing on the application to
transcriptomic analysis. Here we provide transcriptomic case studies
demonstrating how scale models, rather than traditional normalizations,
can reduce false positive and false negative rates in practice while
enhancing the transparency and reproducibility of analyses. We show that
these scale models replace the need for dual cutoff approaches often
used to address the disconnect between practical and statistical
significance. We demonstrate the utility of that scale models built
based on known housekeeping genes in complex metatranscriptomic
datasets. Thus this work provides example and practical guidance on how
to incorporate scale into transcriptomic analysis.
\end{abstract}

@@

Biological systems are both predictably variable and stochastic (46) and
systems biology experiments show that there are transcripts with
approximately constant concentrations in the cell and those with large
variability under different growth conditions (20). Current measurement
methods that rely on high throughput sequencing fail to capture all of
the variation, particularly variation due to scale (3, 16). In the
absence of external information (4, 19, 52) sequencing depth
normalisation methods cannot recapture the scale information (19), and
can only normalize for the technical variation due to sequencing depth.
Here we demonstrated that even approximate estimates of scale and the
uncertainty of measuring it can aid in the interpretation of
RNA-sequencing experiments.

Nixon et al. (3) introduced the idea of explicitly modelling the scale
of a HTS dataset, and showed how to incorporate these models in the
analysis of microbiome and other datasets (16). They demonstrated that
many tools commonly used to analyze HTS datasets had substantial Type 1
and Type 2 error rates, in line with recent findings by others (9). A
version of ALDEx2 with the ability to include scale uncertainty was
shown to be able to correct for the high Type 1 error rate for that
tool, albeit with some loss of sensitivity. Finally, they showed that
incorporating an informed scale model incorporating both location and
scale uncertainty estimates could both control for Type 1 and Type 2
error rates (16).

Building and using a scale model, thus has substantial benefits relative
to the dual cutoff approach that is advocated for many gene expression
experiments (10, 12). In particular, the dual cutoff approach has long
been known to not control for Type 1 errors (13, 14), and the frequent
lack of concordance between tools when benchmarked on transcriptomes(7,
9, 10, 26, 53) and microbiomes (6, 8, 28, 37, 54, 55) suggests poor
control of Type 2 errors as well (7, 9). Thus, incorporating a scale
model during the analysis of HTS data promises the best of both worlds.
A default scale model can control for Type 1 errors with minimal prior
knowledge of the environment and this can be done with essentially no
additional computational overhead. Furthermore, this work and previous
(16) show that even minimal information about the underlying environment
can be used to build a relatively robust scale model that controls for
both Type 1 and 2 error rates.

In the analysis of HTS data it is often observed that larger datasets
converge on the majority of parts being significantly different (3, 9,
10). Li et al. (9) conducted a permutation-based benchmarking study and
found that widely used tools performed worse than simple Wilcoxon
rank-sum tests in controlling the FDR when sample sizes became large.
They suggested that the presence of outliers were one of the factors
driving this observation. Brooks et al. (56) suggested that
inappropriate choice of benchmarking methods are also a major
contributing factor and that objective standards of truth are important.
From the perspective of our work the disagreement between tools can be
explained by the observation that different analytic approaches produce
different parameter estimates. Thus, more data produces worse estimates
because the additional data simply increases the precision of a flawed
estimate (3, 57).

Scale simulation is now build into ALDEx2 (16) and this work suggest
that there are two main root causes to common HTS data pathologies. The
first contributing factor is the observed very low dispersion estimate
for many features that is a by-product of experimental design,
sequencing and normalization. In the Schurch et al. (10) dataset, the
data were from single colonies derived from a single culture. Thus, it
is more accurate to describe the 96 samples as wet-lab technical
replicates rather than independent samples. This type of replication
approach is standard in the molecular literature, and would be expected
to result in the very low dispersion that is observed. Applying the
default scale model with \(\gamma=0.5\) a large number of transcripts
have their dispersion increased (Figure 1D), with the effect being
largest for those with the lowest initial dispersion (Figure 2). Adding
scale uncertainty results in modest number of transcripts, 205, being
called significantly different as shown in the volcano plot in Figure 1B
(red points). In addition, there is now a strong concordance between the
difference and p-values. In hindsight, what is not obvious is why the
unscaled volcano plot shows such poor correspondence. We suggest that
this can be explained by random fluctuations in the many variance
estimates with very low values, and this is supported by the plots shown
in Figure 2.

The second contributing factor is unacknowledged asymmetry in many
datasets (15); i.e., different gene content or a directional change in
the majority of features. In the case of asymmetry, the use of a
user-specified scale model can be very useful for otherwise
difficult-to-analyze datasets such as meta-transcriptomes and in-vitro
selection datasets where the majority of features can change. We showed
one such example for the metatranscriptome dataset in Figure 3. Here the
dataset was highly asymmetrical. Incorporating differential scale on a
per-group basis moves the mass of the housekeeping functions towards the
midline of no difference and so affects both Type I and Type II error
rates. We showed two ways of estimating the scale difference between
groups and found that any reasonable estimate is an improvement over the
default scale model. This is in line with the observations by Nixon et
al (16) in a 16S rRNA gene sequencing dataset. It is also of note that
in the case of true biological replicates (different individuals) that
adding a modest amount of scale \(\gamma=0.5\) had little effect on the
the difference between groups and on the dispersion. Thus, in this
dataset the scale mis-specification was affecting mainly the location of
the difference between groups. While we acknowledge that some prior
information on which housekeeping transcripts should not be classed as
differentially abundant is needed, we suggest that this information is
widely available and is already used when performing the gold-standard
quantitative PCR test of differential abundance (58, 59).

Beyond concerns of fidelity and rigor, scale models also enhance the
reproducibility and transparency of HTS analyses. The addition of scale
uncertainty essentially tests the model over a range of normalizations
(3) and so can replace the consensus approach that has been proposed by
some groups (8, 60) with no additional computational overhead. Thus, an
advantage of incorporating scale is that analyses can be made much more
robust such that actual or potential differences in scale can be tested
and accounted for explicitly. While it is beyond the scope of the
present article, we note that there are many ways of building scale
models that enhance the interpretability of the parameters and
assumptions and a detailed description of these points is describe
elsewhere (3).

In summary, we supply a toolkit that makes incorporating scale
uncertainty and location information simple to incorporate for
transcriptomes or indeed any type of HTS dataset. While the underlying
scale of the system is generally inaccessible, the effect of scale on
the analysis outcomes can be modelled and can help explain some of the
underlying biology, and help to expose known issues with the analysis of
HTS data. Adding scale information to the analysis allows for more
robust inference because the features that are sensitive to scale can be
identified and their impact on conclusions weighted accordingly.
Additionally, the use of informed scale models permits difficult to
analyze datasets to be examined in a robust and principled manner even
when the majority of features are asymmetrically distributed or
expressed (or both) in the groups (47). Thus, using and reporting scale
uncertainty should become a standard practice in the analysis of HTS
datasets.

\singlespacing

@@

\phantomsection\label{refs}
\begin{CSLReferences}{1}{1}
\bibitem[\citeproctext]{ref-lovell:2011}
1. Lovell,D., Müller,W., Taylor,J., Zwart,A. and Helliwell,C. (2011)
Proportions, percentages, ppm: Do the molecular biosciences treat
compositional data right? In Pawlowsky-Glahn,V., Buccianti,A. (eds),
\emph{Compositional Data Analysis: Theory and Applications}. John Wiley;
Sons New York, NY, London, pp. 193--207.

\bibitem[\citeproctext]{ref-Quinn:2019aa}
2. Quinn,T.P., Erb,I., Gloor,G., Notredame,C., Richardson,M.F. and
Crowley,T.M. (2019) \href{https://doi.org/10.1093/gigascience/giz107}{A
field guide for the compositional analysis of any-omics data}.
\emph{Gigascience}, \textbf{8}.

\bibitem[\citeproctext]{ref-nixon2023scale}
3. Nixon,M.P., Letourneau,J., David,L.A., Lazar,N.A., Mukherjee,S. and
Silverman,J.D. (2023) \href{https://arxiv.org/abs/2201.03616}{Scale
reliant inference}. \emph{https://arxiv.org/abs/2201.03616}.

\bibitem[\citeproctext]{ref-Vandeputte:2017aa}
4. Vandeputte,D., Kathagen,G., D'hoe,K., Vieira-Silva,S.,
Valles-Colomer,M., Sabino,J., Wang,J., Tito,R.Y., De Commer,L.,
Darzi,Y., \emph{et al.} (2017)
\href{https://doi.org/10.1038/nature24460}{Quantitative microbiome
profiling links gut community variation to microbial load}.
\emph{Nature}, \textbf{551}, 507--511.

\bibitem[\citeproctext]{ref-Props:2017aa}
5. Props,R., Kerckhof,F.-M., Rubbens,P., De Vrieze,J., Hernandez
Sanabria,E., Waegeman,W., Monsieurs,P., Hammes,F. and Boon,N. (2017)
\href{https://doi.org/10.1038/ismej.2016.117}{Absolute quantification of
microbial taxon abundances}. \emph{ISME J}, \textbf{11}, 584--587.

\bibitem[\citeproctext]{ref-Thorsen:2016aa}
6. Thorsen,J., Brejnrod,A., Mortensen,M., Rasmussen,M.A., Stokholm,J.,
Al-Soud,W.A., Sørensen,S., Bisgaard,H. and Waage,J. (2016)
\href{https://doi.org/10.1186/s40168-016-0208-8}{Large-scale
benchmarking reveals false discoveries and count transformation
sensitivity in 16{S} r{RNA} gene amplicon data analysis methods used in
microbiome studies}. \emph{Microbiome}, \textbf{4}, 62.

\bibitem[\citeproctext]{ref-Quinn:2018aa}
7. Quinn,T.P., Crowley,T.M. and Richardson,M.F. (2018)
\href{https://doi.org/10.1186/s12859-018-2261-8}{Benchmarking
differential expression analysis tools for RNA-seq: Normalization-based
vs. Log-ratio transformation-based methods}. \emph{BMC Bioinformatics},
\textbf{19}, 274.

\bibitem[\citeproctext]{ref-Nearing:2022aa}
8. Nearing,J.T., Douglas,G.M., Hayes,M.G., MacDonald,J., Desai,D.K.,
Allward,N., Jones,C.M.A., Wright,R.J., Dhanani,A.S., Comeau,A.M.,
\emph{et al.} (2022)
\href{https://doi.org/10.1038/s41467-022-28034-z}{Microbiome
differential abundance methods produce different results across 38
datasets}. \emph{Nat Commun}, \textbf{13}, 342.

\bibitem[\citeproctext]{ref-Li:2022aa}
9. Li,Y., Ge,X., Peng,F., Li,W. and Li,J.J. (2022)
\href{https://doi.org/10.1186/s13059-022-02648-4}{Exaggerated false
positives by popular differential expression methods when analyzing
human population samples}. \emph{Genome Biol}, \textbf{23}, 79.

\bibitem[\citeproctext]{ref-Schurch:2016aa}
10. Schurch,N.J., Schofield,P., Gierliński,M., Cole,C., Sherstnev,A.,
Singh,V., Wrobel,N., Gharbi,K., Simpson,G.G., Owen-Hughes,T., \emph{et
al.} (2016) \href{https://doi.org/10.1261/rna.053959.115}{How many
biological replicates are needed in an RNA-seq experiment and which
differential expression tool should you use?} \emph{RNA}, \textbf{22},
839--51.

\bibitem[\citeproctext]{ref-storey2003positive}
11. Storey,J.D. (2003) The positive false discovery rate: A bayesian
interpretation and the q-value. \emph{The annals of statistics},
\textbf{31}, 2013--2035.

\bibitem[\citeproctext]{ref-Cui:2003aa}
12. Cui,X. and Churchill,G.A. (2003)
\href{https://www.ncbi.nlm.nih.gov/pubmed/12702200}{Statistical tests
for differential expression in cDNA microarray experiments}.
\emph{Genome Biol}, \textbf{4}, 210.1--210.10.
